\huge \bf {Using Optimal Time Step Selection to \\ Boost the Accuracy of FD Schemes for \\ Variable-Coefficient PDEs }

### 2017-07-22 papers/2009-OTSPDE-WCE/OTSPDE-WCE2009.tex
Rename papers/OTSPDE-WCE2009 to papers/2009-OTSPDE-WCE

% >>>>>>>>>>>>>>>>>>>>>>>  Put your title here <<<<<<<<<<<<<<<<<<<<<<<<
% make title bold and 14 pt font (Latex default is non-bold, 16pt)
\title{\huge \bf {Using Optimal Time Step Selection to \\ 
       Boost the Accuracy of FD Schemes for \\
       Variable Coefficient PDEs }}

@@

% >>>>>>>>>>>>>>>>>>>>>>>>> Keywords and Abstract <<<<<<<<<<<<<<<<<<<<<
{\hspace{1pc} {\it{\small Abstract}}{\bf{\small---We extend the technique 
of optimal time step (OTS) selection for finite difference (FD) schemes of time 
dependent PDEs to PDEs where the leading-order spatial derivative term has a
spatially varying coefficient.  The basic approach involves identifying a 
transformation of the domain that eliminates the spatial dependence of the 
coefficient for the leading-order term.  This change of variables is then used 
to define an optimal computational grid for the FD scheme on the 
\emph{original} domain.   By using both the optimal grid and OTS selection, we are
able to boost the order of accuracy above what would be expected from a formal
analysis of the FD scheme.  We illustrate the utility of our method by 
applying it to variable coefficient wave and diffusion equations.  In 
addition, we demonstrate the viability of OTS selection for the two-step 
Kreiss-Petersson-Ystr\"om discretization of the wave equation.

\em Keywords: optimal time step selection, optimal grid selection, finite 
    difference schemes, time dependent PDEs, variable coefficient PDEs}}
 }

@@

This paper is organized as follows.  First, we briefly review OTS selection 
for PDEs with a constant-coefficient leading-order spatial derivative term.  
Next, we show how to combine optimal grid and optimal time step selection to 
boost the accuracy of finite difference schemes for variable-coefficient PDEs.  
We then demonstrate the use of variable-coefficient OTS selection on the 
variable-coefficient wave and heat equations.  Our analysis of the wave is 
equation is particularly noteworthy because it is \emph{not} based on a one-step 
time integration scheme.  We conclude with a summary of our main results
and thoughts on future directions for research.

@@

\section{OTS Selection for Variable Coefficient PDEs
  \label{sec:ots_var_coef_pdes}}
Transformation of the spatial domain is a natural way to extend OTS selection 
to semilinear PDEs that have variable coefficient leading-order spatial 
derivatives.  Consider a semilinear PDE of the form:
\bea
  \frac{\partial u}{\partial t} = a^n(x) \frac{\partial^n u}{\partial x^n}
 + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}}, \ldots,
           \frac{\partial u}{\partial x}, u \right)
 + f(x,t)
\label{eq:semilinear_PDE}
\eea
on the domain $0 < x < 1$.  Using the change of variables\footnote{This
transformation is a generalization of the change of variables proposed
in~\cite{guidotti2006} for the wave equation.}
\beq
  y = \Phi(x) = \abar \int_0^x \frac{1}{a(\xi)} d\xi
  \label{eq:change_of_variables}
\eeq
with 
\beq
  \abar \equiv \left( \int_0^1 \frac{1}{a(\xi)} d\xi \right)^{-1},
  \label{eq:abar}
\eeq
we can completely eliminate the spatial dependence of the coefficient on the 
leading-order spatial derivative (at the expense of adding 
lower-order variable-coefficient spatial derivative terms).  The PDE 
(\ref{eq:semilinear_PDE}) on the transformed domain has the form
\bea
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
 + \tilde{F} \left( \frac{\partial^{n-1} u}{\partial y^{n-1}},
           \ldots,
           \frac{\partial u}{\partial y}, u \right)
 + f(y,t),
\label{eq:semilinear_PDE_transformed}
\eea
where $\tilde{F}$ includes any lower-order terms introduced by the process of 
changing variables.

@@

\subsection{Variable Coefficient Wave Equation}
In this section, we apply OTS and optimal grid selection to the
variable coefficient wave equation.  Because we choose a FD scheme based on 
the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
equation~\cite{kreiss2002}, we begin by analyzing OTS selection for 
the constant-coefficient wave equation and demonstrate its applicability to
this multistep method.

@@

\subsubsection*{OTS Selection for Variable Coefficient Wave Equation}
We now consider the variable coefficient second-order wave equation
\beq
    u_{tt} - c(x)^2 u_{xx} = f,
    \label{eq:var_coef_wave_eqn}
\eeq
on the domain $0 < x < 1$.  Using the change of variables defined in 
(\ref{eq:change_of_variables}) and (\ref{eq:abar}), we can transform 
(\ref{eq:var_coef_wave_eqn}) into the constant-coefficient wave equation
\beq
  u_{tt} = \cbar^2 u_{yy} - \cbar c' u_y + f,
\eeq
where $c' = dc/dx$.

\paragraph*{OTS Selection on Transformed Domain}
On the transformed domain, it is straightforward to apply OTS selection 
because the leading-order spatial derivative has a constant coefficient.  
If we use second-order central differences for both the Laplacian and gradient 
terms (on a uniform grid in the transformed domain), the optimal time step is 
$\dt_{opt} = \dx/\cbar$ and the correction term, which is calculated by 
identifying the terms in $\left(\dt^2 \tu_{tttt}/12\right)$ that are not 
eliminated by the use of $\dt_{opt}$, is given by
\bea
  \frac{\dt^2}{12} 
  \left(
    \begin{array}{l}
        -2 \cbar^3 (c')_y u_{yy}
        -\cbar^3 (c')_{yy} u_{y}
        + \cbar^2 (c')^2 u_{yy} 
        \\
        +\ \cbar^2 c' (c')_y u_{y}
        + \cbar^2 f_{yy}
        - \cbar c' f_y
        + f_{tt}
    \end{array}
  \right).
\eea
Using the optimal time step and the correction term, we obtain a fourth-order
accurate KPY scheme for the variable coefficient wave equation 
(see Figure~\ref{fig:var_coef_wave_eqn_1d_error}).  As mentioned earlier,
a fifth-order accurate first time step is required to achieve this level of
accuracy.

\begin{figure}[thb]
\begin{center}
\scalebox{0.30}{\includegraphics{figures/var_coef_wave_eqn_1d_error_vs_N}}
\caption{$L^\infty$ error as a function of number of grid points for various
KPY discretizations of the variable coefficient second-order wave equation:
KPY using a uniform grid on the original domain (red squares),
KPY using a uniform grid on the transformed domain (green triangles),
KPY using the optimal grid without OTS selection the original domain 
(magenta triangles),
KPY using a uniform grid with OTS selection on the transformed domain 
(cyan diamonds),
KPY using the optimal grid with OTS selection on the original domain 
(blue circles).  Note that the green triangles overlap the magenta triangles 
and the cyan diamonds overlap the blue circles. 
}
\label{fig:var_coef_wave_eqn_1d_error}
\end{center}
\end{figure}

@@

As shown in Figure~\ref{fig:var_coef_wave_eqn_1d_error}, by using the optimal 
grid, optimal time step, and the correction terms, we are able to obtain a 
fourth-order accurate scheme for the variable coefficient wave equation on
the original domain.  The error on the variable-spaced grid is almost 
identical to the error obtained when solving the equation on the transformed 
domain.

@@

\subsection{Application to Variable Coefficient Diffusion Equation}
In this section, we apply OTS and optimal grid selection to the 
variable-coefficient diffusion equation
\beq
  u_t = (D(x) u)_{xx} + f(x,t),
  \label{eq:var_coef_diffusion_eqn}
\eeq
on the domain $0 < x < 1$.  Note that we have expressed diffusion in a 
spatially inhomogeneous medium using the Fokker-Planck diffusivity 
law~\cite{vanmilligen2005}, which has been shown to often be more physically 
correct than Fick's law.

Following the procedure outlined in Section~\ref{sec:ots_var_coef_pdes}, the
change of variables used to define the optimal grid is given by
$y = \dbar\int_0^x D(\xi)^{-1/2} d\xi$ where
$\dbar = \left(\int_0^1 D(\xi)^{-1/2} d\xi \right) ^ {-1}$.
Since the optimal time step for the constant coefficient diffusion equation 
is given by $\dto = \dx^2/6D$, the optimal time step for the variable
coefficient diffusion equation is given by $\dto = \left(\dy/\dbar\right)^2/6$.
The correction term for this scheme is 
\bea
  \frac{\dt^2}{2} \left[
  \begin{array}{c}
  6 D D_{xx} u_{xx} + 4 D D_{xxx} u_x D D_{xxxx} u + D f_{xx} \\
  + \ 6 D_x D_{xx} u_x + 2 D_x D_{xxx} u + 2 D_x f_x \\
  + \ D_{xx} u_t + f_t
  \end{array}
  \right]
\eea
which is derived by computing $u_{tt}$ and using the heuristic described in 
Section~\ref{sec:derivation_ots_corr_terms} to exclude the following terms:
\beq
  D^2u_{xxxx} \ , \ 4 D D_x u_{xxx} \ , \ 
  2 D_x D u_{xxx} \ , \ 6 \left(D_x\right)^2 u_{xx}.
\eeq
Combining the optimal grid, optimal time step, and the correction term, we are 
able to obtain fourth-order solutions for (\ref{eq:var_coef_diffusion_eqn})
using only a formally second-order FD scheme (see 
Figure~\ref{fig:var_coef_diffusion_eqn_1d_error}).  
Figure~\ref{fig:var_coef_diffusion_eqn_1d_solns} shows a comparison of 
solutions computed with and without the optimal grid and OTS selection.

\begin{figure}[tbh]
\begin{center}
\scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_OTS_OptGrid_soln}} \ 
\scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_soln}}
\caption{Comparison of numerical solutions for the variable coefficient
diffusion equation~(\ref{eq:var_coef_diffusion_eqn}) computed using 
the optimal grid and OTS selection (left) and a uniform grid without OTS
selection (right).  Both figures compare the analytical solution (solid line)
with the numerical solution (circles) computed using $25$ grid points.
}
\label{fig:var_coef_diffusion_eqn_1d_solns}
\end{center}
\end{figure}

\section{Summary and Conclusions}
In this paper, we have extended the philosophy of optimizing the parameters
of FD schemes (\eg the time step size) for accuracy to the selection of the 
computational grid.  In particular, we have presented a method to boost the 
order of accuracy of FD schemes for PDEs where the leading-order spatial derivative has a 
variable coefficient by using an optimal time step and an optimal grid.  We have 
described the basic procedure for constructing the optimal grid, computing the optimal time step, 
and deriving the required correction terms.  To demonstrate the power of these 
techniques, we have applied them to the variable coefficient wave equation and 
the Fokker-Planck form of the diffusion equation in an inhomogeneous medium.  
Our analysis of the KPY scheme (used to solve the wave equation) also serves 
as an example of OTS selection applied to multistep time integration schemes.

>>>>

% >>>>>>>>>>>>>>>>>>>>>>>  Put your title here <<<<<<<<<<<<<<<<<<<<<<<<
% make title bold and 14 pt font (Latex default is non-bold, 16pt)
\title{\huge \bf {Using Optimal Time Step Selection to \\ 
       Boost the Accuracy of FD Schemes for \\
       Variable-Coefficient PDEs }}

@@

% >>>>>>>>>>>>>>>>>>>>>>>>> Keywords and Abstract <<<<<<<<<<<<<<<<<<<<<
{\hspace{1pc} {\it{\small Abstract}}{\bf{\small---We extend the technique 
of optimal time step (OTS) selection for finite difference (FD) schemes of time 
dependent PDEs to PDEs where the leading-order spatial derivative term has a
spatially varying coefficient.  The basic approach involves identifying a 
transformation of the domain that eliminates the spatial dependence of the 
coefficient for the leading-order term.  This change of variables is then used 
to define an optimal computational grid for the FD scheme on the 
\emph{original} domain.   By using both the optimal grid and OTS selection, we are
able to boost the order of accuracy above what would be expected from a formal
analysis of the FD scheme.  We illustrate the utility of our method by 
applying it to variable-coefficient wave and diffusion equations.  In 
addition, we demonstrate the viability of OTS selection for the two-step 
Kreiss-Petersson-Ystr\"om discretization of the wave equation.

\em Keywords: optimal time step selection, optimal grid selection, finite 
    difference schemes, time dependent PDEs, variable-coefficient PDEs}}
 }

@@

This paper is organized as follows.  First, we briefly review OTS selection 
for PDEs with a constant-coefficient leading-order spatial derivative term.  
Next, we show how to combine optimal grid and optimal time step selection to 
boost the accuracy of finite difference schemes for variable-coefficient PDEs.  
We then demonstrate the use of variable-coefficient OTS selection on the 
variable-coefficient wave and diffusion equations.  Our analysis of the wave is 
equation is particularly noteworthy because it is \emph{not} based on a one-step 
time integration scheme.  We conclude with a summary of our main results
and thoughts on future directions for research.

@@

\section{OTS Selection for Variable-Coefficient PDEs
  \label{sec:ots_var_coef_pdes}}
Transformation of the spatial domain is a natural way to extend OTS selection 
to semilinear PDEs that have variable-coefficient leading-order spatial 
derivatives.  Consider a semilinear PDE of the form:
\bea
  \frac{\partial u}{\partial t} = a^n(x) \frac{\partial^n u}{\partial x^n}
 + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}}, \ldots,
           \frac{\partial u}{\partial x}, u \right)
 + f(x,t)
\label{eq:semilinear_PDE}
\eea
on the domain $0 < x < 1$.  Using the change of variables\footnote{This
transformation is a generalization of the change of variables proposed
in~\cite{guidotti2006} for the wave equation.}
\beq
  y = \Phi(x) = \abar \int_0^x \frac{1}{a(\xi)} d\xi
  \label{eq:change_of_variables}
\eeq
with 
\beq
  \abar \equiv \left( \int_0^1 \frac{1}{a(\xi)} d\xi \right)^{-1},
  \label{eq:abar}
\eeq
we can completely eliminate the spatial dependence of the coefficient on the 
leading-order spatial derivative (at the expense of adding 
lower-order variable-coefficient spatial derivative terms).  The PDE 
(\ref{eq:semilinear_PDE}) on the transformed domain has the form
\bea
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
 + \tilde{F} \left( \frac{\partial^{n-1} u}{\partial y^{n-1}},
           \ldots,
           \frac{\partial u}{\partial y}, u \right)
 + f(y,t),
\label{eq:semilinear_PDE_transformed}
\eea
where $\tilde{F}$ includes any lower-order terms introduced by the process of 
changing variables.

@@

\subsection{Variable-Coefficient Wave Equation}
In this section, we apply OTS and optimal grid selection to the
variable-coefficient wave equation.  Because we choose a FD scheme based on 
the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
equation~\cite{kreiss2002}, we begin by analyzing OTS selection for 
the constant-coefficient wave equation and demonstrate its applicability to
this multistep method.

@@

\subsubsection*{OTS Selection for Variable-Coefficient Wave Equation}
We now consider the variable-coefficient second-order wave equation
\beq
    u_{tt} - c(x)^2 u_{xx} = f,
    \label{eq:var_coef_wave_eqn}
\eeq
on the domain $0 < x < 1$.  Using the change of variables defined in 
(\ref{eq:change_of_variables}) and (\ref{eq:abar}), we can transform 
(\ref{eq:var_coef_wave_eqn}) into the constant-coefficient wave equation
\beq
  u_{tt} = \cbar^2 u_{yy} - \cbar c' u_y + f,
\eeq
where $c' = dc/dx$.

\paragraph*{OTS Selection on Transformed Domain}
On the transformed domain, it is straightforward to apply OTS selection 
because the leading-order spatial derivative has a constant coefficient.  
If we use second-order central differences for both the Laplacian and gradient 
terms (on a uniform grid in the transformed domain), the optimal time step is 
$\dt_{opt} = \dx/\cbar$ and the correction term, which is calculated by 
identifying the terms in $\left(\dt^2 \tu_{tttt}/12\right)$ that are not 
eliminated by the use of $\dt_{opt}$, is given by
\bea
  \frac{\dt^2}{12} 
  \left(
    \begin{array}{l}
        -2 \cbar^3 (c')_y u_{yy}
        -\cbar^3 (c')_{yy} u_{y}
        + \cbar^2 (c')^2 u_{yy} 
        \\
        +\ \cbar^2 c' (c')_y u_{y}
        + \cbar^2 f_{yy}
        - \cbar c' f_y
        + f_{tt}
    \end{array}
  \right).
\eea
Using the optimal time step and the correction term, we obtain a fourth-order
accurate KPY scheme for the variable-coefficient wave equation 
(see Figure~\ref{fig:var_coef_wave_eqn_1d_error}).  As mentioned earlier,
a fifth-order accurate first time step is required to achieve this level of
accuracy.

\begin{figure}[thb]
\begin{center}
\scalebox{0.30}{\includegraphics{figures/var_coef_wave_eqn_1d_error_vs_N}}
\caption{$L^\infty$ error as a function of number of grid points for various
KPY discretizations of the variable-coefficient second-order wave equation:
KPY using a uniform grid on the original domain (red squares),
KPY using a uniform grid on the transformed domain (green triangles),
KPY using the optimal grid without OTS selection the original domain 
(magenta triangles),
KPY using a uniform grid with OTS selection on the transformed domain 
(cyan diamonds),
KPY using the optimal grid with OTS selection on the original domain 
(blue circles).  Note that the green triangles overlap the magenta triangles 
and the cyan diamonds overlap the blue circles. 
}
\label{fig:var_coef_wave_eqn_1d_error}
\end{center}
\end{figure}

@@

As shown in Figure~\ref{fig:var_coef_wave_eqn_1d_error}, by using the optimal 
grid, optimal time step, and the correction terms, we are able to obtain a 
fourth-order accurate scheme for the variable-coefficient wave equation on
the original domain.  The error on the variable-spaced grid is almost 
identical to the error obtained when solving the equation on the transformed 
domain.

@@

\subsection{Application to Variable-Coefficient Diffusion Equation}
In this section, we apply OTS and optimal grid selection to the 
variable-coefficient diffusion equation
\beq
  u_t = (D(x) u)_{xx} + f(x,t),
  \label{eq:var_coef_diffusion_eqn}
\eeq
on the domain $0 < x < 1$.  Note that we have expressed diffusion in a 
spatially inhomogeneous medium using the Fokker-Planck diffusivity 
law~\cite{vanmilligen2005}, which has been shown to often be more physically 
correct than Fick's law.

Following the procedure outlined in Section~\ref{sec:ots_var_coef_pdes}, the
change of variables used to define the optimal grid is given by
$y = \dbar\int_0^x D(\xi)^{-1/2} d\xi$ where
$\dbar = \left(\int_0^1 D(\xi)^{-1/2} d\xi \right) ^ {-1}$.
Since the optimal time step for the constant coefficient diffusion equation 
is given by $\dto = \dx^2/6D$, the optimal time step for the 
variable-coefficient diffusion equation is given by 
$\dto = \left(\dy/\dbar\right)^2/6$.
The correction term for this scheme is 
\bea
  \frac{\dt^2}{2} \left[
  \begin{array}{c}
  6 D D_{xx} u_{xx} + 4 D D_{xxx} u_x + D D_{xxxx} u + D f_{xx} \\
  + \ 6 D_x D_{xx} u_x + 2 D_x D_{xxx} u + 2 D_x f_x \\
  + \ D_{xx} u_t + f_t
  \end{array}
  \right]
\eea
which is derived by computing $u_{tt}$ and using the heuristic described in 
Section~\ref{sec:derivation_ots_corr_terms} to exclude the following terms:
\beq
  D^2u_{xxxx} \ , \ 4 D D_x u_{xxx} \ , \ 
  2 D_x D u_{xxx} \ , \ 6 \left(D_x\right)^2 u_{xx}.
\eeq
Combining the optimal grid, optimal time step, and the correction term, we are 
able to obtain fourth-order solutions for (\ref{eq:var_coef_diffusion_eqn})
using only a formally second-order FD scheme (see 
Figure~\ref{fig:var_coef_diffusion_eqn_1d_error}).  
Figure~\ref{fig:var_coef_diffusion_eqn_1d_solns} shows a comparison of 
solutions computed with and without the optimal grid and OTS selection.

\begin{figure}[tbh]
\begin{center}
\scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_OTS_OptGrid_soln}} \ 
\scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_soln}}
\caption{Comparison of numerical solutions for the variable-coefficient
diffusion equation~(\ref{eq:var_coef_diffusion_eqn}) computed using 
the optimal grid and OTS selection (left) and a uniform grid without OTS
selection (right).  Both figures compare the analytical solution (solid line)
with the numerical solution (circles) computed using $25$ grid points.
}
\label{fig:var_coef_diffusion_eqn_1d_solns}
\end{center}
\end{figure}

\section{Summary and Conclusions}
In this paper, we have extended the philosophy of optimizing the parameters
of FD schemes (\eg the time step size) for accuracy to the selection of the 
computational grid.  In particular, we have presented a method to boost the 
order of accuracy of FD schemes for PDEs where the leading-order spatial 
derivative has a variable coefficient by using an optimal time step and an 
optimal grid.  We have described the basic procedure for constructing the 
optimal grid, computing the optimal time step, and deriving the required 
correction terms.  To demonstrate the power of these techniques, we have 
applied them to the variable-coefficient wave equation and the Fokker-Planck 
form of the diffusion equation in an inhomogeneous medium.  Our analysis of 
the KPY scheme (used to solve the wave equation) also serves as an example of 
OTS selection applied to multistep time integration schemes.
